[![Travis-CI Build Status](https://travis-ci.org/arendsee/rmonad.svg?branch=master)](https://travis-ci.org/arendsee/rmonad)
[![Coverage Status](https://img.shields.io/codecov/c/github/arendsee/rmonad/master.svg)](https://codecov.io/github/arendsee/rmonad?branch=master)

---
output: markdown_github 
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, echo = FALSE}
set.seed(210)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# rmonad

`rmonad` offers

 * nuanced error handling

 * access to the intermediate results of a pipeline

 * effects -- e.g. plotting, caching -- within a pipeline

 * access to results preceding an error

 * tools for branching and combining pipelines

 * a way to almost completely avoid naming things

 * a structured approach to literate programming

## Installation

You can install rmonad from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("arendsee/rmonad")
```

## Examples

For details, see the vignette. Here are a few excerpts

```{r}
library(rmonad)
```

```{r}
1:5      %>>%
    sqrt %v>% # record an intermediate value
    sqrt %>>%
    sqrt
```

```{r}
# Both plots and summarizes an input table
cars %>_% plot(xlab="index", ylab="value") %>>% summary
```

```{r}
x <- list()

# return first value in a list, otherwise return NULL
if(length(x) > 0) {
    x[[1]]
} else {
    NULL
}

# this does the same
x[[1]] %||% NULL %>% esc
```

```{r}
# chain independent pipelines together, with documentation
runif(10)  %>>% abs %>% doc(

    "Alternatively, the documentation could go into a text block below the code
in a knitr document. The advantage of having documentation here, is that it is
coupled unambiguously to the generating function. This is a monadic
interpretation of literate programming. These annotations, together with the
ability to chain chains of monads, allows whole complex workflows to be built,
with the results collated into a single monad. All errors propagate exactly as
errors should, only affecting downstream computations. The final monad can be
converted into a markdown document. A graph of functions can automatically be
built. Summaries of the locations of errors. The monad could be extended for
automated benchmarking."

                  ) %>^% sum %^__%
rnorm("a") %>>% abs %>^% sum %^__%
rexp(10)   %>>% abs %>^% sum %>%
    unbranch
```
