---
title: "Case Study: GFF processing"
author: "Zebulun Arendsee"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case Study: GFF processing}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This work is funded by the National Science Foundation grant [NSF-IOS
1546858](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1546858).


## The problem

Gene Feature Format (GFF) is used to annotate intervals on a genome. Loading
and validating a GFF is a common first step in a bioinformatics analysis.
Mistakes at this step can cause major problems later, so it is important to
validate the GFF and report good diagnostics when problems arise. Here I will
show how such a pipeline can be written using standard methods, and then show
how `rmonad` can be used to organize and annotate such a pipeline.

GFF files are TAB-delimited where each row corresponds to a single interval.
These intervals, though, may be ontologically related. For example a gene is
the parent of an mRNA, which in turn is a parent of a set of exons and a set of
coding sequences (CDS). These relations are specified in the attribute column
(column 9). Here is an example (with TABs replaced with space) that introduces
the main format and the variants we need to support: 

```
# this is a comment, they can appear anywhere in the GFF
# also note, empty lines can appear anywhere in the file

# This is a simple mono-exonic gene
I   .   gene   11565   11951   .   -   .   ID=gene0
I   .   mRNA   11565   11951   .   -   .   ID=mrna0;Parent=gene0
I   .   exon   11565   11951   .   -   .   Parent=mrna0
I   .   CDS    11565   11951   .   -   0   Parent=mrna0

# this is a gene with two splicing variants:
I   .   gene   61931   83591   .   +   .   ID=gene1

I   .   mRNA   61931   83591   .   +   .   ID=rna1;Parent=gene1
I   .   exon   61931   62344   .   +   .   Parent=rna1
I   .   exon   81616   82209   .   +   .   Parent=rna1
I   .   exon   82211   83591   .   +   .   Parent=rna1
I   .   CDS   61931   62344   .   +   0   Parent=rna1
I   .   CDS   81616   82209   .   +   0   Parent=rna1
I   .   CDS   82211   82681   .   +   0   Parent=rna1

I   .   mRNA   61931   83591   .   +   .   ID=rna2;Parent=gene1
I   .   exon   61931   62344   .   +   .   Parent=rna2
I   .   exon   81616   82209   .   +   .   Parent=rna2
I   .   exon   82211   83591   .   +   .   Parent=rna2
I   .   CDS   61931   62344   .   +   0   Parent=rna2
I   .   CDS   82211   82681   .   +   0   Parent=rna2

# Below are a few variants that occur (unforunately) in the wild

# V1: CDS directly descending from a gene.
b   .   gene   7235   9016   .   -   .   ID=gene2
b   .   CDS    7235   9016   .   -   0   Parent=gene2

# V2: 'Parent=-' when feature has no parent
I   .   gene   11565   11951   .   -   .   ID=gene3;Parent=-

# V3: Elements with no tags that need to be treated as IDs
I   .   gene   11565   11951   .   -   .   gene3
I   .   mRNA   11565   11951   .   -   .   ID=mrna3;Parent=gene3
I   .   exon   11565   11951   .   -   .   Parent=mrna3
I   .   CDS    11565   11951   .   -   0   Parent=mrna3
```

Another issues we need to account for is type synonyms. The feature type
(column 3) is required to be valid Sequence Ontology (SO) terms. For the
purposes of this vignette, I will just handle the following sets of synonyms:

 * gene := `gene` | `SO:0000704`

 * mRNA := `mRNA` | `messenger RNA` | `messenger_RNA` | `SO:0000234`
            | `transcript` | `SO:0000673`

 * CDS  := `CDS` | `coding_sequence` | `coding sequence` | `SO:0000316`

 * exon := `exon` | `SO:0000147` | `coding_exon` | `coding exon` |`SO:0000195` 

For mRNA and exon, I am merging to ontology terms (mRNA and transcript; exon
and coding exon). Formally, this is incorrect, but pratically it is probably
the right thing. Since these transformations may be wrong, they need to be
noted.

## Solution

To test our solution, I use the `gff` rmonad dataset.

### Conventional approach

Before using `rmonad`, I will use a more conventional approach.

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(magrittr)
set.seed(210)

good_gfffile <- system.file('extdata', 'gff', '0.gff3', package='rmonad')

read_gff <- function(file){
  readr::read_tsv(
    file,
    col_names = c(
        "seqid",
        "source",
        "type",
        "start",
        "stop",
        "score",
        "strand",
        "phase",
        "attr"
    ),
    na        = ".",
    comment   = "#",
    col_types = "ccciidcic"
  )
}

read_gff(good_gfffile)
```

This is a nice start, also, `readr` will pick up on any deviations from the
specified column number and type, warning of problems:

```{r}
# wrong type
bad0 <- system.file('extdata', 'gff', 'bad-0.gff3', package='rmonad')
bad_result0 <- read_gff(bad0)
bad_result0
```

Notice that the 4th entry has a start value of "NA". This indicates there was something wrong with he gff file and when we take a look at the original gff, we see that the entry corresponds to the string "one thousand eight hundred and seven" instead of the number. While this is not a likely error, it is a similar error for when a number contains a comma (example: "1,807"") which R will parse as a string.

```{,eval=FALSE}
# incorrect type
a       Strex   region  1       230218  .       +       .       ID=I
a       Strex   telomere        1       801     .       -       .       ID=id1
a       Strex   origin_of_replication   707     776     .       +       .       ID=id2
a       Strex   gene    "one thousand eight hundred and seven"  2169    .       -       .       ID=gene1
```

Similarly when we load in a gff missing a column.
```{r}
# wrong column number
bad1 <- system.file('extdata', 'gff', 'bad-1.gff3', package='rmonad')
bad_result1 <- read_gff(bad1)
bad_result1
```
The read_gff can correctly identify that a column is missing from the following raw file.
```{,eval=FALSE}
# missing one column
a       region  1       230218  .       +       .       ID=I
a       telomere        1       801     .       -       .       ID=id1
a       origin_of_replication   707     776     .       +       .       ID=id2
a       gene    1807    2169    .       -       .       ID=gene1
```

R uses `NA` to indicate missing values. The R 'numeric' type corresponds to
Haskell `[Maybe Num]`, i.e. an array of possibly empty values. In a GFF,
columns 2,6,7,8 and 9 may be missing, the others may not. So we need an
additional assertion that these are complete.

```{r}
g <- read_gff(good_gfffile)
for(col in c("seqid", "type", "start", "stop")){
  if(any(is.na(g[[col]]))){
    stop("GFFError: Column '", col, "' may not have missing values")
  }
}
```

Now we can account for type synonyms using the following method

```{r}
gene_synonyms <- 'SO:0000704'
mRNA_synonyms <- c('messenger_RNA', 'messenger RNA', 'SO:0000234')
CDS_synonyms  <- c('coding_sequence', 'coding sequence', 'SO:0000316')
exon_synonyms <- 'SO:0000147'

g$type <- ifelse(g$type %in% gene_synonyms, 'gene', g$type)
g$type <- ifelse(g$type %in% mRNA_synonyms, 'mRNA', g$type)
g$type <- ifelse(g$type %in% CDS_synonyms,  'CDS',  g$type)
g$type <- ifelse(g$type %in% exon_synonyms, 'exon', g$type)

mRNA_near_synonyms <- c('transcript', 'SO:0000673')
exon_near_synonyms <- c('SO:0000147', 'coding_exon', 'coding exon', 'SO:0000195')

if(any(g$type %in% mRNA_near_synonyms)){
    g$type <- ifelse(g$type %in% mRNA_near_synonyms, 'mRNA', g$type)
    warning("Substituting transcript types for mRNA types, this is probably OK")
}

if(any(g$type %in% exon_near_synonyms)){
    g$type <- ifelse(g$type %in% exon_near_synonyms, 'exon', g$type)
    warning("Substituting transcript types for exon types, this is probably OK")
}
```

Now we need to evaluate the attribute column.

```{r}

tags <- c("ID", "Parent")

data_frame(
    attr  = stringr::str_split(g$attr, ";"),
    order = 1:nrow(g)
  ) %>%
  dplyr::mutate(ntags = sapply(attr, length)) %>%
  tidyr::unnest(attr) %>%
  dplyr::mutate(attr = ifelse(grepl('=', attr), attr, paste(".U", attr, sep="="))) %>%
  tidyr::separate_(
    col   = "attr",
    into  = c("tag", "value"),
    sep   = "=",
    extra = "merge"
  ) %>%
  dplyr::filter(tag %in% c(tags, ".U")) %>%
  {
    if(nrow(.) > 0){
      tidyr::spread(., key="tag", value="value")
    } else {
      .$tag   = NULL
      .$value = NULL
      .
    }
  } %>%
  {
    if("Parent" %in% names(.)){
      .$Parent <- ifelse(.$Parent == "-", NA, .$Parent)
    }
    .
  } %>% {
    for(tag in c(tags, ".U")){
      if(! tag %in% names(.))
        .[[tag]] = NA_character_
    }
    .
  } %>%
  {
    if("ID" %in% names(.))
      .$ID <- ifelse(is.na(.$ID) & !is.na(.$.U) & .$ntags == 1, .$.U, .$ID)
    .
  } %>%
  merge(data_frame(order=1:nrow(g)), all=TRUE) %>%
  dplyr::arrange(order) %>%
  { cbind(g, .) } %>%
  dplyr::select(-.U, -order, -ntags, -attr) %>%
  {
    if(all(c("ID", "Parent") %in% names(.))){
      parents <- subset(., type %in% c("CDS", "exon"))$Parent
      parent_types <- subset(., ID %in% parents)$type

      if(any(parent_types == "gene"))
        warning("Found CDS or exon directly inheriting from a gene, this may be fine.") 

      if(! all(parent_types %in% c("gene", "mRNA")))
        stop("Found CDS or exon with illegal parent")

      if( any(is.na(parents)) )
        stop("Found CDS or exon with no parent")

      if(! any(duplicated(.$ID, incomparables=NA)))
        warning("IDs are not unique, this is probably bad")
    }
    .
  }

```

The beauty of this chain is that it requires few temporary variables (just `g`,
and `tags`), it is a pure flow of data. It is an elegant sequence of functions
operating on a single thread of data.

But there are a few problems.

First, it is in dire need of documentation. We could add in comments. But
comments cannot be formatted well. A better approach is some form of literate
programming, such as rewriting the program in Rmarkdown. But this 1) breaks the
pipeline (since we can't pipe between chunks), 2) results in an object we can't
compute on, 3) makes debugging even more difficult, because our code is spread
out. 

### `rmonad` approach

With `rmonad` we can mingle documentation and code in a computable object.

```{r}
require(readr)
require(dplyr)
require(tidyr)
require(magrittr)
library(rmonad)
set.seed(210)

#' @param file The input GFF file
#' @param tags The GFF tags to keep
#'        default these are not stored)
read_gff <- function(file, tags){

  "
  This function reads a GFF. It builds a data frame from it that with a column
  for each value in `tags`.

  The `read_gff` function is a nested function, containing an Rmonad pipeline
  inside it. This docstring describes the entire wrapped pipeline.
  "

  raw_gff <- as_monad(
    {

      "
      Rmonad supports docstrings. If a block begins with a string, this
      string is extracted and stored. Python has something similar, where the
      first string in a function is cast as documentation.
      
      The `as_monad` function takes an expression and wraps its result into a
      context. It also handles the extraction of this docstring. The result
      here is used at more than one place in the pipeline. Rather than
      accessing it later as a global, it will be funneled bach in.
      "

      readr::read_tsv(
        file,
        col_names = c(
          "seqid",
          "source",
          "type",
          "start",
          "stop",
          "score",
          "strand",
          "phase",
          "attr"
        ),
        na        = ".",
        comment   = "#",
        col_types = "ccciidcic"
      )
    }
  )

  raw_gff %>% tag('raw_gff') %>>% {

    "
    The %>>% operator applies the function described in this block to the
    input on the left-hand-side. This corresponds to the UNIX '|' or magrittr's
    '%>%'. It differs from them in that it is a monadic bind operator, rather
    than an application operator. It carries a context along with the
    computations. The context can store past values, performance information,
    this docstring, and links to the parent chunk. The context is a directed
    graph of code chunks and their metadata.

    Here I assert that no data is missing in the columns where it is required.
    "

    for(col in c("seqid", "type", "start", "stop")){
      if(any(is.na(.[[col]]))){
        stop("GFFError: Column '", col, "' may not have missing values")
      }
    }
    .
  } %>>% {

    "
    According the GFF specification, the value in the type column should be
    from the Sequence Ontology (SO). Here I collapse the synonymous terms
    for the main features usually present in GFF file (gene, mRNA, CDS, and
    exon). In production code, I would probably do something a bit more formal. 
    "

    gene_synonyms <- 'SO:0000704'
    mRNA_synonyms <- c('messenger_RNA', 'messenger RNA', 'SO:0000234')
    CDS_synonyms  <- c('coding_sequence', 'coding sequence', 'SO:0000316')
    exon_synonyms <- 'SO:0000147'

    .$type <- ifelse(.$type %in% gene_synonyms, 'gene', .$type)
    .$type <- ifelse(.$type %in% mRNA_synonyms, 'mRNA', .$type)
    .$type <- ifelse(.$type %in% CDS_synonyms,  'CDS',  .$type)
    .$type <- ifelse(.$type %in% exon_synonyms, 'exon', .$type)

    .
  } %>_% {

    "
    The %>_% operator lets this chunk of code be run for its effects, which are
    raising warnings if we replace the type with a questionable synonym.  We
    could alternatively just use %>>% and add a terminal '.' to this chunk.
    However, the `%>_%` operator guarantees the input is passed without being
    changed, which is safer and also clarifies our intention. 
    "

    mRNA_near_synonyms <- c('transcript', 'SO:0000673')
    exon_near_synonyms <- c('SO:0000147', 'coding_exon', 'coding exon', 'SO:0000195')

    if(any(.$type %in% mRNA_near_synonyms)){
        .$type <- ifelse(.$type %in% mRNA_near_synonyms, 'mRNA', .$type)
        warning("Substituting transcript types for mRNA types, this is probably OK")
    }

    if(any(.$type %in% exon_near_synonyms)){
        .$type <- ifelse(.$type %in% exon_near_synonyms, 'exon', .$type)
        warning("Substituting transcript types for exon types, this is probably OK")
    }

  } %>>% {

    "
    Notice below that I use the magrittr operator '%>%' inside the rmonad
    pipeline. When to pipe with rmonad and when to pipe with magrittr is a
    matter of granularity. This chunk of code perhaps should form one
    documentation unit. And perhaps I don't expect it to fail. If I break this
    chunk into several, the failures are more localized, and I can access
    intermediate values for debugging. On the other hand, putting every little
    operation in a new chunk will clutter the graph and reports derived from
    it.
    "

    data_frame(
      attr  = stringr::str_split(.$attr, ";"),
      order = 1:nrow(.)
    ) %>%
      dplyr::mutate(ntags = sapply(attr, length)) %>%
      tidyr::unnest(attr) %>%
      dplyr::mutate(attr = ifelse(grepl('=', attr), attr, paste(".U", attr, sep="="))) %>%
      tidyr::separate_(
        col   = "attr",
        into  = c("tag", "value"),
        sep   = "=",
        extra = "merge"
      )

   } %>% funnel(raw_gff=raw_gff, tags=tags) %*>% {

    "
    The %v>% operator stores the input value in memory. We could replace every
    %>>% operator with %v>%. This would let us inspect every step of an
    analysis at the cost of high memory usage. For brevity, I won't break this
    following block down any further.

    The `funnel` function packages a list in a monad, merging their histories
    and propagating error. That is, if `gff` or `tags` failed upstream, this
    function will not be run. `%*>%` takes a list on the left and feeds it into
    the function on the right as an argument list. Here `funnel` and `%*>%` are
    used together to merge a pipeline (gff) and inject a parameter (tags).

    We could not have written

      %v>% function(gff=gff, tags=tags)

    because this would have brought the monad wrapped gff into scope, not the
    value itself.
    "

      dplyr::filter(., tag %in% c(tags, ".U")) %>%
      {
        if(nrow(.) > 0){
          tidyr::spread(., key="tag", value="value")
        } else {
          .$tag   = NULL
          .$value = NULL
          .
        }
      } %>%
      {
        if("Parent" %in% names(.)){
          .$Parent <- ifelse(.$Parent == "-", NA, .$Parent)
        }
        .
      } %>% {
        for(tag in c(tags, ".U")){
          if(! tag %in% names(.))
            .[[tag]] = NA_character_
        }
        .
      } %>%
      {
        if("ID" %in% names(.))
          .$ID <- ifelse(is.na(.$ID) & !is.na(.$.U) & .$ntags == 1, .$.U, .$ID)
        .
      } %>%
      merge(data_frame(order=1:nrow(raw_gff)), all=TRUE) %>%
      dplyr::arrange(order) %>%
      { cbind(raw_gff, .) } %>%
      dplyr::select(-.U, -order, -ntags, -attr)

  } %>_% {

    "
    And make the last few assertions.
    "

    if(all(c("ID", "Parent") %in% names(.))){
      parents <- subset(., type %in% c("CDS", "exon"))$Parent
      parent_types <- subset(., ID %in% parents)$type

      if(any(parent_types == "gene"))
        warning("Found CDS or exon directly inheriting from a gene, this may be fine.")

      if(! all(parent_types %in% c("gene", "mRNA")))
        stop("Found CDS or exon with illegal parent")

      if(any(is.na(parents)) )
        stop("Found CDS or exon with no parent")

      if(any(duplicated(.$ID, incomparables=NA)))
        warning("IDs are not unique, this is probably bad")
    }

  } %>_% {

    "
    I could post some closing comments here. The %>_% operator can be chained
    and the output does not affect the output of the main chain. The NULL is
    required to distinguish this block from an anonymous function that returns
    a string.
    "

    NULL
  }
  # End Rmonad chain

}

gff_as_graph <- function(file){
  read_gff(file, tags=c("ID", "Parent")) %>>%
  dplyr::filter(!is.na(Parent)) %>>%
  {
      "Make a unique label for each CDS and exon interval"
      dplyr::group_by(., Parent) %>%
      dplyr::mutate(ID = ifelse(type %in% c("CDS", "exon"),
                                paste(type, start, sep='-'),
                                ID)) %>%
      dplyr::ungroup()
  } %>>% {
    "Make a graph of gene models"
    dplyr::select(., ID, Parent) %>%
      as.matrix %>%
      igraph::graph_from_edgelist()
  }
}

```

That is the whole GFF program in an rmonad framework

```{r}
result <-
  system.file('extdata', 'gff', '0.gff3', package='rmonad') %v>% gff_as_graph
result %>% plot
```

`esc` will extract the final result, and raise all errors, warnings and
messages that where extracted

```{r}
esc(result)
```

Now we see why we might want a little more granularity in our pipeline. To
summarize the results we can use the `mtabulate` functions.

```{r}
mtabulate(result)
```

We can also get a summary of issues

```{r}
missues(result)
```

The `id` column corresponds to a row number in the `mtabulate` result.

To extract particular values, we can use the `get_*` family of vectorized
getters

```{r, eval=FALSE}
# get a list of every stored value, report uncached values as NULL
get_value(result, warn=FALSE)
# get a list of every docstring
get_doc(result)
```

```{r gff-workflow-plot, fig.width=6, fig.height=6}
plot(result)
```

A plot of the connections between functions in the pipeline. The node labels
are the `rmonad` ids. The red arrow indicates a 'nest' relationship. Green nodes
are OK, orange nodes raised a warning, red nodes (none appear in this graph)
represent errors.

If the pipeline fails, the last valid result is saved, for example

```{r}
m <-
  system.file('extdata', 'gff', '15.gff3', package='rmonad') %v>%
  gff_as_graph
```

To diagnose the problem, we might first read the error messages:

```{r}
get_error(m)
```

Here we see the 4th node raised an error: we having missing values in the
'type' column.

To get a big picture of the pipeline:

```{r}
mtabulate(m)
```

The the last three nodes are in the failing state (see column 'OK'). We can see
all issues:

```{r}
missues(m)
```

We see two warnings in node 3. To see what code is wrapped in node 3:

```{r}
get_code(m, 3)
```

So there were issues in reading the file. We can get the value of this node's parent:

```{r}
get_value(m, get_parents(m, 3)[[1]])[[1]]
```

If we wanted to explore the raw data, we could jump into a Bash shell, navigate
to the file above, and take it apart with UNIX tools (the lines of raw data
were not stored in the Rmonad object).

We might next check out the failing columns:

```{r}
get_code(m, 4:6)
```

Now that we know exactly _where_ the problem is, we can start working through
the data that triggered the problem:

```{r}
get_value(m, get_parents(m, 5)[[1]])
get_value(m, 5)[[1]]
```


From this it looks like there is raw sequence data contained in the GFF file.
According to the GFF specification, it is legal to have the full sequence data
appended to the tabular GFF data. So we probably should either add a test that
will raise an appropriate error message or extend our function to handle this
case.

`rmonad` saved the input to the failing function. This makes debugging much
simpler. We did not have to disassemble our pipeline and rerun it step-by-step.

Overall, in `rmonad`, the output of a pipeline is not just the effluent of the
last node, but the collection of all the nodes along the way. The pipeline
itself becomes data that can be computed upon. 

```{r, echo=FALSE}
# # save data, this is used in testing and examples
# gff <- list(good_result=result, bad_result0 = bad_result0, bad_result1 = bad_result1)
# save(gff, file="gff.rda", version=3)
```
