---
title: "Rmonad: an introduction"
author: "Zebulun Arendsee"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE, message=FALSE}
library(rmonad)
library(magrittr)
set.seed(210)
```

`Rmonad` offers

 * a stateful pipeline framework

 * nuanced error handling

 * access to the intermediate results of a pipeline

 * effects -- e.g. plotting, caching -- within a pipeline

 * access to results preceding an error

 * branching and chaining of pipelines

 * a structured approach to literate programming


## Monadic pipelines

I will introduce `Rmonad` with a simple sequence of squares

```{r}
# %>>% corresponds to Haskell's >>=
1:5      %>>%
    sqrt %>>%
    sqrt %>>%
    sqrt
```

So what exactly did Rmonad do with your data? It is still there, sitting
happily inside the monad.

In `magrittr` you could do something similar:

```{r}
1:5      %>%
    sqrt %>%
    sqrt %>%
    sqrt
```

`%>%` takes the value on the left and applies it to the function on the right.
`%>>%`, takes a monad on the left and a function on the right, then builds a
new monad from them. This new monad holds the computed value, if the
computation succeeded. It collates all errors, warnings, and messages.  These
are stored in step-by-step a history of the pipeline.

`%>%` is an application operator, `%>>%` is a *monadic bind* operator. Magrittr
and Rmonad complement eachother. `%>%` can be used inside a monadic sequence to
perform operations *on* monads, whereas `%>>%` performs operations *in* them.
If this is all too mystical, just hold on, the examples are sensical even
without an understanding of monads.

Below, we store an intermediate value in the monad:

```{r}
1:5      %>>%
    sqrt %v>% # store this result
    sqrt %>>%
    sqrt
```

The `%v>%` variant of the *monadic bind* operator stores the results as they
are passed.

Warnings are caught and stored

```{r}
-1:3     %>>%
    sqrt %v>%
    sqrt %>>%
    sqrt
```

Similarly for errors

```{r}
"wrench" %>>%
    sqrt %v>%
    sqrt %>>%
    sqrt
```

The first `sqrt` failed, and this step was coupled to the resultant error.
Contrast this with `magrittr`, where the location of the error is lost:

```{r, error=TRUE}
"wrench" %>%
    sqrt %>%
    sqrt %>%
    sqrt
```

Also note that a value was still produced. This value will never be used in the
downstream monadic sequence (except when explicitly doing error handling).
However it, and all other information in the monad, can be easily accessed. 



## Extracting data from an Rmonad

If you want to extract the terminal result from the monad, you can use the `esc`
function:

```{r}
1:5 %>>% sqrt %>% esc
```

`esc` is our first example of a class of functions that work on monads, rather
than the values they wrap. We use `magrittr`'s application operator `%>%` here,
rather than the monadic bind operator `%>>%`, because we are passing a literal
monad to `esc`.

If the monad is in a failed state, `esc` will raise an error.

```{r, error=TRUE}
"wrench" %>>% sqrt %>>% sqrt %>% esc
```

If you prefer a tabular summary of your results, you can pipe the monad into
the `mtabulate` function.

```{r}
1:5      %>>%
    sqrt %v>%
    sqrt %>>%
    sqrt %>% mtabulate
```

The `id` colum holds a unique id for each step in the monadic sequence. These
can be extracted with the `extract_by_id`. This function is given such a long
awkward name because it is not stable yet, and will soon be replaced with
something better.

All errors, warnings and notes can be extracted with the `missues` command

```{r}
-2:2 %>>% sqrt %>>% colSums %>% missues
```


## Handling effects

The `%>_%` operator is useful when you want to include a function inside a
pipeline that should be bypassed, but you want the errors, warnings, and
messages to pass along with the main.

You can cache an intermediate result

```{r, eval=FALSE}
cars %>_% write.csv(file="cars.tab") %>>% summary
```

Or plot a value along with a summary

```{r}
cars %>_% plot(xlab="index", ylab="value") %>>% summary %>% forget
```

I pipe the final monad into `forget`, which is (like `esc`) a function for
operating on monads. `forget` removes history from a monad. I do this just to
de-clutter the output.

You can call multiple effects

```{r, eval=FALSE}
cars                                 %>_%
    plot(xlab="index", ylab="value") %>_%
    write.csv(file="cars.tab")       %>>%
    summary %>% forget
```

Since state is passed,  assertions about the data inside a pipeline.

```{r}
iris                                    %>_%
    { stopifnot(is.data.frame(.))     } %>_%
    { stopifnot(sapply(.,is.numeric)) } %>>%
    colSums %|>% head
```

The above code will enter a failed state if the input is either not a data
frame or the columns are not all numeric. The braced expressions are anonymous
functions of `.` (as in `magrittr`). The final expression `%|>%` catches an
error and performs `head` on the last valid input (`iris`).


## Error handling

Errors needn't be viewed as abnormal. For example, we might want to try several
alternatives functions, and use the first that works. 

```{r}
1:10 %>>% colSums %|>% sum
```

Here we will do either `colSums` or `sum`. The pipeline fails only if both
fail.

Sometimes you want to ignore the previous failure completely, and make a new
call -- for example in reading files:

```{r, eval=FALSE}
# try to load a cached file, on failure rerun the analysis
read.table("analyasis_cache.tab") %||% run_analysis(x)
```

This can also be used to replace if-else if-else strings

```{r}
x <- list()
# compare
if(length(x) > 0) { x[[1]] } else { NULL }
# to 
x[[1]] %||% NULL %>% esc
```

Or maybe you want to support multiple extensions for an input file

```{r, eval=FALSE}
read.table("a.tab") %||% read.table("a.tsv") %>>% dostuff
```

Used together with `%|>%` we can build full error handling pipelines

```{r}
letters[1:10] %v>% colSums %|>% sum %||% message("Can't process this")
```


Overall, in Rmonad, errors are well-behaved. It is reasonable to write
functions that return an error rather than one of the myriad default values
(`NULL`, `NA`, `logical(0)`, `list()`, `FALSE`). This is unambiguous. Rmonad
can catch the error and allow allow the programmer to deal with it accordingly.



## Branching pipelines

If you want to perform an operation on a value inside the chain, but don't want
to pass it, you can use the branch operator `%>^%`.

```{r}
rnorm(30) %>^% qplot(xlab="index", ylab="value") %>>% mean %>% unbranch
```

This stores the result of `qplot` in a branch off the main pipeline. This means
that `plot` could fail, but the rest of the pipeline could continue. You can
store multiple branches.

```{r}
rnorm(30) %>^% plot(xlab="index", ylab="value") %>^% summary %>>% mean
```

Branches can be used as input, as well.

```{r}
x <- 1:10 %>^% dgamma(10, 1) %>^% dgamma(10, 5) %^>% cor
x
unbranch(x)
```

Note the branches could be long monadic chains themselves, which might have
their own branches. The `unbranch` function recursively extracts all branches
from the tree.


## Chains of chains

If you want to connect many chains, all with independent inputs, you can do so
with the `%__%` and `%v__%` operators.

```{r}
runif(10) %>>% sum %v__%
rnorm(10) %>>% sum %v__%
rexp(10)  %>>% sum
```

The `%v__%` operator records the output of the lhs and evaluates the rhs into
an Rmonad. If you don't care about the output, you can use `%__%`, which simply
replaces the rhs value with lhs value (while, of course, propagating state).



## Multiple inputs

So far our pipelines have been limited to either linear paths or the somewhat
awkward branch merging. An easier approach is to read inputs from a list. But
we want to be able to catch errors resulting from evaluation of each member of
the list. We can do this with `list_meval`.

```{r}
lsmeval(
    "yolo",
    stop("stop, drop, and die"),
    runif("simon"),
    k = 2
)
```

This returns a monad which fails if any of the components evaluate to an error.
But it does not toss the rest of the inputs, instead returning a clean list
with a NULL filling in missing pieces. Constrast this with normal list evaluation:

```{r, error=TRUE}
list( "yolo", stop("stop, drop, and die"), runif("simon"), 2)
```

`lsmeval` records each failure in each element of the list independently.

This approach can also be used with the infix operator `%*>%`.

```{r, eval=FALSE}
lsmeval(read.csv("a.csv"), read.csv("b.csv")) %*>% merge
```

Now, of course, we can add monads to the mix 

```{r, eval=FALSE}
lsmeval(
    a = read.csv("a.csv") %>>% do_analysis_a,
    b = read.csv("b.csv") %>>% do_analysis_b,
    k = 5
) %*>% joint_analysis
```

Monadic list evaluation is the natural way to build large programs from smaller
pieces.



## Annotating steps

As our pipelines become more complex, it becomes essential to document them. We
can do that as follows:

```{r}
runif(5) %>>% abs %>% doc(

    "Alternatively, the documentation could go into a text block below the code
    in a knitr document. The advantage of having documentation here, is that it
    is coupled unambiguously to the generating function. This is a monadic
    interpretation of literate programming. These annotations, together with
    the ability to chain chains of monads, allows whole complex workflows to be
    built, with the results collated into a single monad. All errors propagate
    exactly as errors should, only affecting downstream computations. The final
    monad can be converted into a markdown document and automatically generated
    function graphs."

                  ) %>^% sum %__%
rnorm(6)   %>>% abs %>^% sum %v__%
rnorm("a") %>>% abs %>^% sum %__%
rexp(6)    %>>% abs %>^% sum %T>%
  {print(mtabulate(.)) } %>% missues
```

Note `%T>%` is a `magrittr` operator. It allows execution of the rhs function
on the lhs *monad* and bypasses the result.
