---
title: "`rmonad`: an introduction"
author: "Zebulun Arendsee"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This work is funded by the National Science Foundation grant [NSF-IOS
1546858](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1546858).


```{r, echo=FALSE, message=FALSE}
library(rmonad)
library(magrittr)
set.seed(210)
```

`rmonad` offers

 * a stateful pipeline framework

 * pure error handling

 * access to the intermediate results of a pipeline

 * effects -- e.g. plotting, caching -- within a pipeline

 * branching and chaining of pipelines

 * a flexible approach to literate programming



## Monadic pipelines

I will introduce `rmonad` with a simple sequence of squares

```{r}
# %>>% corresponds to Haskell's >>=
1:5      %>>%
    sqrt %>>%
    sqrt %>>%
    sqrt
```

So what exactly did `rmonad` do with your data? It is still there, sitting
happily inside the monad.

In `magrittr` you could do something similar:

```{r}
1:5      %>%
    sqrt %>%
    sqrt %>%
    sqrt
```

`%>%` takes the value on the left and applies it to the function on the right.
`%>>%`, takes a monad on the left and a function on the right, then builds a
new monad from them. This new monad holds the computed value, if the
computation succeeded. It collates all errors, warnings, and messages. These
are stored in step-by-step a history of the pipeline.

`%>%` is an application operator, `%>>%` is a *monadic bind* operator. `magrittr`
and `rmonad` complement eachother. `%>%` can be used inside a monadic sequence to
perform operations *on* monads, whereas `%>>%` performs operations *in* them.
If this is all too mystical, just hold on, the examples are sensical even
without an understanding of monads.

Below, we store an intermediate value in the monad:

```{r}
1:5      %>>%
    sqrt %v>% # store this result
    sqrt %>>%
    sqrt
```

The `%v>%` variant of the *monadic bind* operator stores the results as they
are passed.

Following the example of `magrittr`, arbirary anonymous functions of '.' are
supported

```{r}
1:5 %>>% { o <- . * 2 ; { o + . } %>% { . + o } }
```

Warnings are caught and stored

```{r}
-1:3     %>>%
    sqrt %v>%
    sqrt %>>%
    sqrt
```

Similarly for errors

```{r}
"wrench" %>>%
    sqrt %v>%
    sqrt %>>%
    sqrt
```

The first `sqrt` failed, and this step was coupled to the resultant error.
Contrast this with `magrittr`, where the location of the error is lost:

```{r, error=TRUE}
"wrench" %>%
    sqrt %>%
    sqrt %>%
    sqrt
```

Also note that a value was still produced. This value will never be used in the
downstream monadic sequence (except when explicitly doing error handling).
However it, and all other information in the monad, can be easily accessed. 


## Extracting data from an `rmonad`

If you want to extract the terminal result from the monad, you can use the `esc`
function:

```{r}
1:5 %>>% sqrt %>% esc
```

`esc` is our first example of a class of functions that work on monads, rather
than the values they wrap. We use `magrittr`'s application operator `%>%` here,
rather than the monadic bind operator `%>>%`, because we are passing a literal
monad to `esc`.

If the monad is in a failed state, `esc` will raise an error.

```{r, error=TRUE}
"wrench" %>>% sqrt %>>% sqrt %>% esc
```

If you prefer a tabular summary of your results, you can pipe the monad into
the `mtabulate` function.

```{r}
1:5      %>>%
    sqrt %v>%
    sqrt %>>%
    sqrt %>% mtabulate
```

An internal states can be accessed by converting the monad to a list of past
states and simple indexing out the ones you want.


All errors, warnings and notes can be extracted with the `missues` command

```{r}
-2:2 %>>% sqrt %>>% colSums %>% missues
```

The `id` column refers to row numbers in the `mtabulate` output. Internal
values can be extracted by converting the monad to a list and indexing:

```{r}
result <- 1:5 %v>% sqrt %v>% sqrt %v>% sqrt
as.list(result)[[2]] %>% esc
```

An `rmonad` can be converted to a `DiagrammeR` graph (or `igraph`); in this way,
standard tools for network analysis can be applied to pipelines. `rmonad`
currently has built in support for plotting piplines and making markdown
reports (experimental).

```{r, eval=FALSE}
as_dgr_graph(result)
```

## Handling effects

The `%>_%` operator is useful when you want to include a function inside a
pipeline that should be bypassed, but you want the errors, warnings, and
messages to pass along with the main.

You can cache an intermediate result

```{r, eval=FALSE}
cars %>_% write.csv(file="cars.tab") %>>% summary
```

Or plot a value along with a summary

```{r}
cars %>_% plot(xlab="index", ylab="value") %>>% summary %>% forget
```

I pipe the final monad into `forget`, which is (like `esc`) a function for
operating on monads. `forget` removes history from a monad. I do this just to
de-clutter the output.

You can call multiple effects

```{r, eval=FALSE}
cars                                 %>_%
    plot(xlab="index", ylab="value") %>_%
    write.csv(file="cars.tab")       %>>%
    summary %>% forget
```

Since state is passed, you can make assertions about the data inside a
pipeline.

```{r}
iris                                    %>_%
    { stopifnot(is.data.frame(.))     } %>_%
    { stopifnot(sapply(.,is.numeric)) } %>>%
    colSums %|>% head
```

The above code will enter a failed state if the input is either not a data
frame or the columns are not all numeric. The braced expressions are anonymous
functions of '.' (as in `magrittr`). The final expression `%|>%` catches an
error and performs `head` on the last valid input (`iris`).



## Error handling

Errors needn't be viewed as abnormal. For example, we might want to try several
alternatives functions, and use the first that works. 

```{r}
1:10 %>>% colSums %|>% sum
```

Here we will do either `colSums` or `sum`. The pipeline fails only if both
fail.

Sometimes you want to ignore the previous failure completely, and make a new
call -- for example in reading files:

```{r, eval=FALSE}
# try to load a cached file, on failure rerun the analysis
read.table("analyasis_cache.tab") %||% run_analysis(x)
```

This can also be used to replace if-else if-else strings

```{r}
x <- list()
# compare
if(length(x) > 0) { x[[1]] } else { NULL }
# to 
x[[1]] %||% NULL %>% esc
```

Or maybe you want to support multiple extensions for an input file

```{r, eval=FALSE}
read.table("a.tab") %||% read.table("a.tsv") %>>% dostuff
```

Used together with `%|>%` we can build full error handling pipelines

```{r}
letters[1:10] %v>% colSums %|>% sum %||% message("Can't process this")
```


Overall, in `rmonad`, errors are well-behaved. It is reasonable to write
functions that return an error rather than one of the myriad default values
(`NULL`, `NA`, `logical(0)`, `list()`, `FALSE`). This approach is unambiguous.
`rmonad` can catch the error and allow allow the programmer to deal with it
accordingly.



## Branching pipelines

If you want to perform an operation on a value inside the chain, but don't want
to pass it, you can use the branch operator `%>^%`.

```{r, eval=FALSE}
rnorm(30) %>^% qplot(xlab="index", ylab="value") %>>% mean
```

This stores the result of `qplot` in a branch off the main pipeline. This means
that `plot` could fail, but the rest of the pipeline could continue. You can
store multiple branches.

```{r, eval=FALSE}
rnorm(30) %>^% qplot(xlab="index", ylab="value") %>^% summary %>>% mean
```

Branches can be used as input, as well.

```{r}
x <- 1:10 %>^% dgamma(10, 1) %>^% dgamma(10, 5) %^>% cor
x
unbranch(x)
```

Note the branches could be long monadic chains themselves, which might have
their own branches. The `unbranch` function recursively extracts all branches
from the tree.



## Chains of chains

If you want to connect many chains, all with independent inputs, you can do so
with the `%__%` operator.

```{r}
runif(10) %>>% sum %__%
rnorm(10) %>>% sum %__%
rexp(10)  %>>% sum
```

The `%__%` operator records the output of the lhs and evaluates the rhs into an
`rmonad`. This operator is a little like a semicolon, in that it demarcates
independent statements. Each statement, though, is wrapped into a graph of
operations. This graph is itself data, and can be computed on. You could take
any analysis and recompose it as `%__%` delimited blocks. The result of running
the analysis would be a data structure containing all results and errors.

```{r, eval=FALSE}
program <-
{
    x = 2
    y = 5
    x * y
} %__% {
    letters %>% sqrt
} %__% {
    10 * x
}
```

You can link chunks of code, with their results, and performance information.



## Multiple inputs

So far our pipelines have been limited to either linear paths or the somewhat
awkward branch merging. An easier approach is to read inputs from a list. But
we want to be able to catch errors resulting from evaluation of each member of
the list. We can do this with `list_meval`.

```{r}
funnel(
    "yolo",
    stop("stop, drop, and die"),
    runif("simon"),
    k = 2
)
```

This returns a monad which fails if any of the components evaluate to an error.
But it does not toss the rest of the inputs, instead returning a clean list
with a NULL filling in missing pieces. Constrast this with normal list evaluation:

```{r, error=TRUE}
list( "yolo", stop("stop, drop, and die"), runif("simon"), 2)
```

`funnel` records each failure in each element of the list independently.

This approach can also be used with the infix operator `%*>%`.

```{r, eval=FALSE}
funnel(read.csv("a.csv"), read.csv("b.csv")) %*>% merge
```

Now, of course, we can add monads to the mix 

```{r, eval=FALSE}
funnel(
    a = read.csv("a.csv") %>>% do_analysis_a,
    b = read.csv("b.csv") %>>% do_analysis_b,
    k = 5
) %*>% joint_analysis
```

Monadic list evaluation is the natural way to build large programs from smaller
pieces.


## Annotating steps

As our pipelines become more complex, it becomes essential to document them. We
can do that as follows:

```{r}
{

    "This is docstring. The following list is metadata associated with this
    node. Both the docstring and the metadata list will be processed out of
    this function before it is executed. They also will not appear in the code
    stored in the Rmonad object."

    list(sys = sessionInfo(), foo = "This can be anything")

    # This NULL is necessary, otherwise the metadata list above would be
    # treated as the node output
    NULL

} %__% # The %__% operator connects independent pieces of a pipeline.

"a" %>>% {

    "The docstrings are stored in the Rmonad objects. They may be extracted in
    the generation of reports. For example, they could go into a text block
    below the code in a knitr document. The advantage of having documentation
    here, is that it is coupled unambiguously to the generating function. These
    annotations, together with the ability to chain chains of monads, allows
    whole complex workflows to be built, with the results collated into a
    single object. All errors propagate exactly as errors should, only
    affecting downstream computations. The final object can be converted into a
    markdown document and automatically generated function graphs."

    paste(., "b")

}
```


## Nesting pipelines

`rmonad` pipelines may be nested to arbitrary depth. 

```{r}
foo <- function(x, y) {
    "This is a function containing a pipeline. It always fails"    

    "a" %>>% paste(x) %>>% paste(y) %>>% log
}

bar <- function(x) {
    "this is another function, it doesn't fail"

    funnel("b", "c") %*>% foo %>>% paste(x)
}

"d" %>>% bar
```

This function descends through three levels of nesting. There is a failure at
the deepest level. This failing node, where a string is passed to a `log`
function, stores the error message and the input. Each node ascending from the
point of failure stores their respective input. This allows debugging to resume
from any desired level.
