---
title: "'rmonad' with 'Nozzle.R1'"
author: "Zebulun Arendsee"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{'rmonad' with 'Nozzle.R1'}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This vignette shows how `rmonad` can be used in tandem with `Nozzle.R1` to
create HTML reports. `Nozzle.R1` is a published R package for report
generation:

```{r}
library(Nozzle.R1)
citation('Nozzle.R1')
```

It can be used to build HTML reports section by section. The package offers a
report generation approach that is more programable, in some ways, than
standard literate programming approaches (e.g. `knitr`). `rmonad` is intended
to store all the data needed for a report, but stop short of actually creating
one. These two packages are thus highly complementary.

Here is a simple example:

```{r, eval=FALSE}
m <- iris %>>% head(20) %v>% summary %v>% colSums

r <- newCustomReport("Demo report")
r <- setMaintainerName(r, "Uriah Heep")

s1 <- newSection("Pipeline overview")

fig1 <- "p1.png"
png(fig1)
plot(m)
dev.off()

s1 <- addTo(s1,
  newFigure(fig1, "Pipeline graph"),
  newTable(mtabulate(m), "Pipeline summary")
)
r <- addTo(r, s1)

s2 <- newSection("Results")
s2 <- addTo(s2, newTable(get_value(m, 2)[[1]], get_code(m, 2)[[1]]))
s2 <- addTo(s2, newTable(get_value(m, 2)[[1]], get_code(m, 2)[[1]]))
r <- addTo(r, s2)

if(any(has_error(m) | has_warnings(m))){ 
    s3 <- newSection("Bad things")
    s3 <- addTo(s3, newTable(missues(m), "Issues raised"))
    r <- addTo(r, s3)
}

writeReport(r, filename="demo")
```
